---
title: "redist: Markov Chain Monte Carlo Methods for Redistricting Simulation"
output: 
  md_document:
    variant: markdown_github
---
# redist: Markov Chain Monte Carlo Methods for Redistricting Simulation [![Build Status](https://travis-ci.org/kosukeimai/redist.svg?branch=master)](https://travis-ci.org/kosukeimai/redist) [![CRAN_Status_Badge](http://www.r-pkg.org/badges/version/redist)](https://cran.r-project.org/package=redist)
Authors:

- [Ben Fifield](https://www.benfifield.com), bfifield@princeton.edu (Maintainer)
- Alex Tarr, atarr@princeton.edu
- [Michael Higgins](http://www-personal.k-state.edu/~mikehiggins/), mjh5@princeton.edu
- [Kosuke Imai](https://imai.princeton.edu), kimai@princeton.edu

Paper: [A New Automated Redistricting Simulator Using Markov Chain Monte Carlo](http://imai.princeton.edu/research/files/redist.pdf)

## Installation Instructions
`redist` is available on CRAN and can be installed using:
```{r eval = FALSE}
install.packages("redist")
```

You can also install the most recent development version of `redist` using the `devtools` package.  First you have to install `devtools` using the following code. Note that you only have to do this once:
```{r eval=FALSE} 
if(!require(devtools)) install.packages("devtools")
```   
Then, load `devtools` and use the function `install_github()` to install `redist`:

```{r eval=FALSE}
library(devtools)
install_github("kosukeimai/redist",dependencies=TRUE)
```

# Usage Examples - 25 Precincts into 3 Districts
## No Population Constraint
```{r eval = FALSE}
## Load data
library(redist)
data(algdat.pfull)

## Run the simulations
mcmc.out <- redist.mcmc(adjobj = algdat.pfull$adjlist,
                        popvec = algdat.pfull$precinct.data$pop,
                        nsims = nsims,
                        ndists = 3)
```

## 20% Population Constraint
```{r eval = FALSE}
## Load data
data(algdat.p20)

## -----------------------------------------------
## Run mcmc algorithm - hard population constraint
## (reject any sample where population parity > 20%)
## -----------------------------------------------
mcmc.out <- redist.mcmc(adjobj = algdat.p20$adjlist,
                        popvec = algdat.p20$precinct.data$pop,
                        nsims = nsims,
                        popcons = .2,
                        ndists = 3)

## ---------------------------------------------------------------------
## Run mcmc algorithm - draws from gibbs defined by distance from parity
## (Run with no tempering)
## ---------------------------------------------------------------------
mcmc.out.gb <- redist.mcmc(adjobj = algdat.p20$adjlist,
                           popvec = algdat.p20$precinct.data$pop,
                           ndists = 3,
                           nsims = nsims,
                           constraint = "population",
                           constraintweights = 5.4)
## Reweight draws back to the uniform distribution
mcmc.out.gb <- redist.ipw(mcmc.out.gb, targetpop = .2)

## ---------------------------------------------------------------------
## Run mcmc algorithm - draws from gibbs defined by distance from parity
## (Run with simulated tempering, betas power law sequence of length 10
## from 0 to 1)
## Also including optional beta weights, to upweight prob of transitions
## to colder temperatures
## ---------------------------------------------------------------------
betaweights <- rep(NA, 10); for(i in 1:10){betaweights[i] <- 2^i}
mcmc.out.st <- redist.mcmc(adjobj = algdat.p20$adjlist,
                           popvec = algdat.p20$precinct.data$pop,
                           ndists = 3,
                           nsims = nsims,
                           constraint = "population",
                           constraintweights = 5.4,
                           temper = TRUE,
                           betaweights = betaweights)
mcmc.out.st <- redist.ipw(mcmc.out.st, targetpop = .2)
```

## 20% Population Constraint and Compactness Constraint
```{r eval = FALSE}
## ----------------------------------------------------------
## Constrain on population and compactness with tempering,
## weight on population = 5.4 while weight on compactness = 2
## Also specifying argument for ssdmat, the distance matrix
## ----------------------------------------------------------
mcmc.out.st.multiple <- redist.mcmc(adjobj = algdat.p20$adjlist,
                                    popvec = algdat.p20$precinct.data$pop,
                                    ndists = 3,
                                    nsims = nsims,
                                    constraint = c("population", "compact"),
                                    constraintweights = c(5.4, 3),
                                    ssdmat = algdat.p20$distancemat,
                                    temper = TRUE,
                                    betaweights = betaweights)
mcmc.out.st <- redist.ipw(mcmc.out.st.multiple, targetpop = .2)
```