---
title: "Map Pre-processing for Special Constraints"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Map Pre-processing for Special Constraints}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
set.seed(5118)
```

In redistricting analysis, it is often useful or necessary to analyze only a
portion of a map, or hold some districts fixed while others are re-simulated.
Other analyses might require a status-quo-type constraint that encourages
simulated districts to be close to a reference plan.

All of these requirements may be achieved through map pre-processing, and this
vignette shows how to use `redist` to do so, using the Metropolitan King County
Council districts as an example.

# The Map
King County is the most populous county in the state of Washington, and contains
the city of Seattle. The nine members of the county council are elected from
single-member districts which are redrawn every decade. According to the 
[county charter](https://aqua.kingcounty.gov/council/clerk/code/03_Charter.htm), 
the districts should be "drawn to produce districts with compact and
contiguous territory, composed of economic and geographic units and
approximately equal in population," and should follow municipality lines as
much as possible.

The precinct data are available online, and contain population, presidential
vote, city, and existing district information.

```{r setup, message=F}
library(dplyr)
library(ggplot2)
library(scales)
library(patchwork)

library(redist)

data_url = "https://github.com/alarm-redist/redist-data/raw/main/data/king_county.rds"
data_path = tempfile()
download.file(data_url, data_path)
king_shp = readRDS(data_path)
print(king_shp)
```

```{r include=F}
n_city = length(unique(king_shp$city)) - 1
tot_area = as.numeric(sum(sf::st_area(king_shp)))
tot_pop = sum(king_shp$pop)
unincorp_area = as.numeric(sum(sf::st_area(king_shp[king_shp$city == "UNINCORP", ])))
unincorp_pop = sum(king_shp$pop[king_shp$city == "UNINCORP"])
```

There are `r n_city` incorporated cities in King County,
which together cover `r percent(1-unincorp_area/tot_area)` of the population 
and `r percent(1-unincorp_area/tot_area)` of the area of the county. The
remainder is "unincorporated King County".  The county contains a significant
amount of water, too, which complicates the drawing of districts; Vashon Island
in the southwest part of the county is not connected to the rest of the county
by land.

We'll start by looking at some maps of the county and the districts.

```{r water-plot, fig.width=8, echo=F}
areas = as.numeric(units::set_units(sf::st_area(king_shp), mi^2))
pop_plot = ggplot(king_shp, aes(fill=pop/areas)) +
    geom_sf(size=0) +
    scale_fill_viridis_c(trans="sqrt", labels=comma, limits=c(0, 25e3), oob=squish) +
    labs(title="Population Density") +
    theme_void() + theme(legend.position="bottom")
water_plot = ggplot(king_shp, aes(fill=pct_water)) +
    geom_sf(size=0) +
    scale_fill_viridis_c(labels=percent) +
    labs(title="Water") +
    theme_void() + theme(legend.position="bottom")
water_plot + pop_plot
```

```{r city-distr-plot, echo=F}
cities = summarize(group_by(king_shp, city, distr), .groups="drop")
districts = summarize(group_by(king_shp, distr))
ggplot(cities) +
    geom_sf(aes(fill=city, alpha=city!="UNINCORP"), color="#444444") +
    geom_sf(color="black", size=1, data=districts, fill="#00000000") +
    geom_sf_text(aes(label=distr), size=10, fontface="bold", color="#000000aa",
                 data=districts) +
    scale_alpha_manual(values=c(0, 0.8)) +
    guides(fill="none", alpha="none") +
    labs(title="Cities with Council District Overlays") + 
    theme_void()
```

# Creating the `redist_map` object
The first step one in a `redist` analysis is creating the `redist_map` object,
which stores the basic parameters of the redistricting problem.  Among these 
parameters is the desired level of population parity.  Here, we'll compute
the parity of the current set of districts, and ensure that all of our
simulations do no worse.

```{r}
existing_parity = redist.parity(king_shp$distr, king_shp$pop)
king = redist_map(king_shp, existing_plan=distr, pop_tol=existing_parity)
print(king)
```

This `redist_map` object contains an adjacency graph for the county. We 
can explore this graph, and zoom in on the city of Seattle, using
`plot()`.

```{r king-adj, fig.height=6}
plot(king, adj=T, centroids=F, zoom_to=(city == "SEA"))
```


# Subsetting
Often, we wish to restrict our analysis to a part of a map or only a few of 
the districts.  This is supported in `redist` using the `filter()` function
from [`dplyr`](https://dplyr.tidyverse.org/).  The package's version of `filter()`
will automatically update the adjacency graph, the number of districts, and
the relevant population bounds.

## Specific districts
Suppose we wanted to study districts 2, 4, and 8, which cover most of Seattle.

```{r}
filter(king, distr %in% c(2, 4, 8))
```

Looking at the information in the header, and comparing it to the original `king` object,
we see that the number of districts has been updated from 9 to 3, and the population
tolerances have been updated from 214,506 ± 1.663% to 213,049 - 0.9909% and
213,049 + 2.358%. Not visible but equally important are the edits to the
adjacency graph to reflect the new geometry of the map.

## Dealing with water and islands
Another way we might want to subset would be to cut out the precincts which are
just water, so that districts won't unnecessarily cross bodies of water. Of
course, we'll have to ensure that Vashon Island is still connected to the
mainland by at least one precinct. We'll start by subsetting to the water
precincts and plotting labels.

```{r king-water}
plot(filter(king, pct_water >= 0.99, pop == 0)) + geom_sf_text(aes(label=id))
```

We see that by removing all water precincts except WVPS34 and WVSP34, we can
maintain a connection between the island and the mainland (incidentally, the
state ferry connecting the island to Seattle runs through these precincts).

```{r king-land}
water_prec = filter(king, pct_water >= 0.99, pop == 0) %>% pull(id)
water_prec = setdiff(water_prec, c("WVPS34", "WVSP34"))
king_land = filter(king, !(id %in% water_prec))
plot(king_land)
```

Zooming in again to view the adjacency graph in the city of Seattle, we see that
the graph has been appropriately edited to remove the water precincts.

```{r seattle-land-adj, fig.height=6}
plot(king_land, adj=T, centroids=F, zoom_to=(city == "SEA"))
```

# Merging
Often, we want to merge some units together to form larger units, either to
visualize or analyze at a coarser scale, or to ensure that the merged units
are treated as one "block" in any redistricting algorithm.  Merging units 
is a part of most map preprocessing workflows, and in `redist` is carried out 
by the `merge_by()` function, which works like a combination of the `group_by()`
and `summarize()` verbs of `dplyr`.

For example, we could merge our King County data by city.

```{r}
merge_by(king_land, city)
```

Under the hood, `merge_by()` does several things. First, it groups the shapefile
by the provided key or keys (here, `city`). By default it also groups by
existing districts, so that the merged units will still follow district
boundaries. Then for each remaining column, `merge_by()` tries to automatically
summarize it.  Most numeric columns are summed (but columns with percentage
values are averaged), and most character columns are collapsed into summary
variables. You can read more about the details of this process in 
[the documentation](../reference/merge_by.html). Finally, `merge_by()` makes
the appropriate edits to the adjacency graph.

Merging geographic shapefile units can be computationally intensive, and so 
by default `merge_by()` drops the geometry before merging.  This is OK for 
analysis purposes, since all of the relevant adjacency information is still
encoded in the graph.  After analysis, you can use the `pullback()` method to
un-merge objects and restore plotting capability.  However, should you want
to preserve the geometry through merging, you can simply set `drop_geom=FALSE`.

```{r merged-city}
king_merged = merge_by(king_land, city, drop_geom=FALSE)
plot(king_merged, adj=T)
```

We will see more uses of `merge_by()` in the sections below.

# Freezing
Sometimes, rather than completely remove a portion of a map, we want to freeze
it in place, so that all of the units in that portion stay together in the same
district. The reasons for doing this might vary, but include enforcing a 
county or administrative boundary split constraint, aiding in setting up Voting
Rights Act constraints, or preparing a map to be optimized according to 
a set of criteria.

In the context of King County Council seats, we might want to implement the
requirement that districts follow municipal lines by ensuring that any sampled
redistricting plans not split any municipalities which are not split by the
existing plan. That way, the number of split municipalities in the set of 
sampled plans will be guaranteed to not exceed the number of existing splits.

In `redist`, freezing is accomplished by using the `freeze()` and `merge_by()`
functions. The former takes in a description of the units which should be
frozen, and groups them into contiguous chunks of frozen units, returning 
an indexing vector that uniquely identifies each group. Then `merge_by()`
merges these groups together.  We can use the `redist.splits()` function to
count split municipalities, and the `is_county_split()` function to identify
split municipalities.

```{r}
cat(redist.splits(king_land$distr, king_land$city), "split cities\n")

king_land %>%
    mutate(is_unsplit = !is_county_split(distr, city)) %>%
plot(is_unsplit)

king_unsplit = king_land %>%
    mutate(unsplit_id = freeze(!is_county_split(distr, city))) %>%
    merge_by(unsplit_id, city, collapse_chr=FALSE)
print(king_unsplit)
```

The plot above shows which cities will be frozen together so that they cannot
be split. Notice that we merge by not just `unsplit_id` but also `city`, so
that adjacent unsplit cities are not merged together. We also set
`collapse_chr=FALSE` to drop the `id` and `precinct` columns, which become
slightly unwieldy after a large merge.

To see this in action, we'll sample 50 redistricting plans using `redist_smc()`
on this partially frozen map.  We'll use `pullback()` to then reconstruct
the plan output of `redist_smc()` so that it is congruous with the original
geometry object, `king_land`.

```{r unsplit-plan, fig.width=8}
plans = redist_smc(king_unsplit, 50, silent=T)
print(plans)
print(pullback(plans))
redist.plot.plans(pullback(plans), draws=1:4, geom=king_land)
```

Notice how the `Merged from another map...` line disappears and the number of
map units changes after using `pullback()`. Notice also that each of the sampled
plans completely preserves the municipalities which were frozen with `freeze()`
and `merge_by()`.

# District Cores
A common requirement in redistricting is that districts after redistricting
resemble the original districts, or 
["preserve the cores" of previous districts](https://www.ncsl.org/research/redistricting/redistricting-criteria.aspx),
to ensure relative continuity of representation. The `redist` package
operationalizes this idea by explicitly constructing the cores of a set of
districts with  `make_cores()`, and merging them together with `merge_by()`.

The idea for constructing district cores is to work inwards from district
boundaries.  First, we merge each district completely.  Then, we un-freeze the
precincts which lie along district boundaries. Then, we un-freeze the precincts
which were unfrozen in the previous step. We repeat this process a user-specified
number of times, leaving only the central "cores" of each district frozen. When
it comes time to simulate, these cores will be assigned a district as a unit,
preserving representation for all the people living in the core.

The `make_cores()` function takes a `boundary` parameter which counts the number
of these steps; `boundary=1` corresponds to un-freezing the precincts along the
boundary only. This is often sufficient for a moderate-to-strong status quo
constraint.

For example, if we set `boundary=1` in King county, 82% of the population lives
inside a district core, while with `boundary=2` that number drops to 58%.

```{r}
pop_inside_cores = function(boundary) {
    king_land %>%
        mutate(core = make_cores(boundary=boundary)) %>%
        group_by(core) %>%
        filter(n() > 2) %>% # filter to cores only
        pull(pop) %>% sum
}
pop_inside_cores(1) / sum(king_land$pop)
pop_inside_cores(2) / sum(king_land$pop)
```

Here, we'll use `boundary=1`. We can see how large areas inside each district
are merged together after applying `merge_by()` to the generated cores.

```{r cores}
king_cores = king_land %>%
    mutate(core = make_cores(boundary=1)) %>%
    merge_by(core, drop_geom=F)
plot(king_cores)
```

If we sample redistrict plans from this modified map, we observe that the
simulated plans generally follow the same location and shape as the original
plan. Sampling is also faster, since there are fewer units in the map.

```{r core-plans}
plans = redist_smc(king_cores, 50, silent=T)
redist.plot.plans(plans, draws=1:4, geom=king_cores)
```

